---
title: "Data Processing for Final Project"
subtitle: "MSK-CHORD 2024 · Breast Cancer · BIOSTAT 629"
author: "Yiqiao Zhu"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Objective

Build the **analytic dataset** (`df_analysis.rds`) for the final project from
three MSK-CHORD 2024 raw files + the curated gene panel. Final dataset contains
all variables needed for the subtype-stratified survival analysis:

- **Outcome**: OS_MONTHS, OS_EVENT
- **Primary exposures**: mutation indicators for all genes in `final_panel`
- **Effect modifier (subtype)**: SUBTYPE (HR × HER2 → LumA / LumB / HER2E / TNBC)
- **Confounders**: AGE, STAGE (I–IV)
- **Sensitivity-analysis variables**: TMB, SAMPLE_TYPE (primary vs metastasis)
- **Descriptive**: RACE

---

# Setup

```{r libraries}
library(dplyr)
library(tidyr)
library(readr)
library(knitr)
```

```{r paths}
# ---- Paths (Mac absolute; edit if moved) ----------------------------------
proj_dir     <- "/Users/yiqiaozhu/Desktop/BIOS 629"
final_dir    <- file.path(proj_dir, "Final Project")
mskchord_dir <- file.path(proj_dir, "msk_chord_2024")
dataset_dir  <- file.path(final_dir, "dataset")
dir.create(dataset_dir, showWarnings = FALSE, recursive = TRUE)

patient_path <- file.path(mskchord_dir, "data_clinical_patient.txt")
sample_path  <- file.path(mskchord_dir, "data_clinical_sample.txt")
mut_path     <- file.path(mskchord_dir, "data_mutations.txt")
panel_path   <- file.path(dataset_dir, "final_panel.rds")
```

---

# Step 1 — Load raw files and curated panel

```{r load-raw}
pat  <- read.delim(patient_path, comment.char = "#",
                   stringsAsFactors = FALSE, check.names = FALSE)
samp <- read.delim(sample_path,  comment.char = "#",
                   stringsAsFactors = FALSE, check.names = FALSE)
mut  <- read.delim(mut_path,     comment.char = "#",
                   stringsAsFactors = FALSE, check.names = FALSE)

final_panel <- readRDS(panel_path)

cat("Patient rows   :", nrow(pat), "\n")
cat("Sample rows    :", nrow(samp), "\n")
cat("Mutation rows  :", nrow(mut), "\n")
cat("Gene panel     :", length(final_panel), "genes\n")
print(final_panel)
```

---

# Step 2 — Restrict to breast cancer cohort

```{r filter-bc}
bc_samp <- samp %>% filter(CANCER_TYPE == "Breast Cancer")
bc_ids  <- unique(bc_samp$PATIENT_ID)

cat("BC samples :", nrow(bc_samp), "\n")
cat("BC patients:", length(bc_ids), "\n")
```

---

# Step 3 — Patient-level mutation indicators (all panel genes)

A patient is classified as mutated (1) for a given gene if **any** of their
tumor samples harbors a mutation in that gene; wild-type (0) otherwise.

```{r mutation-indicators}
bc_mut <- mut %>%
  filter(Tumor_Sample_Barcode %in% bc_samp$SAMPLE_ID,
         Hugo_Symbol %in% final_panel) %>%
  inner_join(bc_samp %>% select(SAMPLE_ID, PATIENT_ID),
             by = c("Tumor_Sample_Barcode" = "SAMPLE_ID"))

mut_wide <- bc_mut %>%
  distinct(PATIENT_ID, Hugo_Symbol) %>%
  mutate(x = 1L) %>%
  pivot_wider(names_from = Hugo_Symbol, values_from = x, values_fill = 0L)

# Ensure every panel gene has a column (even if 0 mutations)
for (g in final_panel) {
  if (!g %in% names(mut_wide)) mut_wide[[g]] <- 0L
}

# Reorder columns: PATIENT_ID first, then genes in panel order
mut_wide <- mut_wide[, c("PATIENT_ID", final_panel)]

cat("Patients with >=1 mutation in panel genes:", nrow(mut_wide), "\n\n")
cat("Mutation rates across panel genes:\n")
sapply(mut_wide[, final_panel], mean) %>%
  round(3) %>% sort(decreasing = TRUE) %>% print()
```

---

# Step 4 — Aggregate sample-level attributes to patient level

Some patients have multiple tumor samples. We create patient-level versions of
STAGE, TMB, and SAMPLE_TYPE using the following rules:

- **STAGE**: take the **highest stage** recorded across samples (worst-case;
  matches the clinical interpretation of cohort staging).
- **TMB**: take the **max TMB** across samples (most-mutated tumor represents
  the genomic burden).
- **SAMPLE_TYPE**: indicator `HAS_PRIMARY` (= 1 if any sample is Primary).
  Useful for the primary-only sensitivity analysis.

```{r recode-stage-helper}
# Map CLINICAL_GROUP string to integer stage 1-4 (NA otherwise).
recode_stage <- function(x) {
  dplyr::case_when(
    x %in% c("1", "1A", "1B")      ~ 1L,
    x %in% c("2A", "2B")            ~ 2L,
    x %in% c("3A", "3B", "3C")      ~ 3L,
    x == "4"                         ~ 4L,
    TRUE                             ~ NA_integer_
  )
}
```

```{r sample-aggregation}
bc_samp2 <- bc_samp %>%
  mutate(STAGE_INT = recode_stage(CLINICAL_GROUP),
         TMB_NUM   = suppressWarnings(as.numeric(TMB_NONSYNONYMOUS)))

sample_agg <- bc_samp2 %>%
  group_by(PATIENT_ID) %>%
  summarise(
    STAGE        = suppressWarnings(max(STAGE_INT, na.rm = TRUE)),
    TMB          = suppressWarnings(max(TMB_NUM,   na.rm = TRUE)),
    HAS_PRIMARY  = as.integer(any(SAMPLE_TYPE == "Primary", na.rm = TRUE)),
    N_SAMPLES    = n(),
    .groups      = "drop"
  ) %>%
  mutate(
    STAGE = ifelse(is.infinite(STAGE), NA_integer_, STAGE),
    TMB   = ifelse(is.infinite(TMB),   NA_real_,    TMB)
  )

cat("Patient-level aggregated sample attributes:\n")
summary(sample_agg %>% select(STAGE, TMB, HAS_PRIMARY, N_SAMPLES))
```

---

# Step 5 — Clean patient-level clinical data

```{r clinical-clean}
pat_bc <- pat %>%
  filter(PATIENT_ID %in% bc_ids) %>%
  select(PATIENT_ID, OS_MONTHS, OS_STATUS,
         CURRENT_AGE_DEID, RACE, HR, HER2) %>%
  rename(AGE = CURRENT_AGE_DEID) %>%
  mutate(
    OS_EVENT = dplyr::case_when(
      OS_STATUS == "1:DECEASED" ~ 1L,
      OS_STATUS == "0:LIVING"   ~ 0L,
      TRUE                      ~ NA_integer_
    ),
    # SUBTYPE from HR x HER2
    SUBTYPE = dplyr::case_when(
      HR == "Yes" & HER2 == "No"  ~ "LumA",
      HR == "Yes" & HER2 == "Yes" ~ "LumB",
      HR == "No"  & HER2 == "Yes" ~ "HER2E",
      HR == "No"  & HER2 == "No"  ~ "TNBC",
      TRUE                         ~ NA_character_
    )
  )

cat("HR distribution:\n");    print(table(pat_bc$HR, useNA = "ifany"))
cat("\nHER2 distribution:\n");  print(table(pat_bc$HER2, useNA = "ifany"))
cat("\nSUBTYPE distribution:\n"); print(table(pat_bc$SUBTYPE, useNA = "ifany"))
```

---

# Step 6 — Merge all components

```{r merge}
df <- pat_bc %>%
  left_join(mut_wide,   by = "PATIENT_ID") %>%
  left_join(sample_agg, by = "PATIENT_ID") %>%
  # patients with no mutation record -> wild-type for all panel genes
  mutate(across(all_of(final_panel), ~ ifelse(is.na(.), 0L, as.integer(.))))

cat("Rows after merge:", nrow(df), "\n")
cat("Unique patients :", length(unique(df$PATIENT_ID)), "\n")
```

---

# Step 7 — Apply inclusion / exclusion filters

Exclude patients with missing values in any <b>key analysis variable</b>:
OS_MONTHS, OS_EVENT, AGE, STAGE, SUBTYPE. Excluded are also Stage 0 (DCIS) and
unknown stage (Stage 9) — these are already mapped to NA in `recode_stage`.

```{r filter}
cat("Missingness BEFORE filter:\n")
sapply(df %>% select(OS_MONTHS, OS_EVENT, AGE, STAGE, SUBTYPE),
       function(x) sum(is.na(x))) %>% print()

df <- df %>%
  filter(!is.na(OS_MONTHS), !is.na(OS_EVENT),
         !is.na(AGE),       !is.na(STAGE),
         !is.na(SUBTYPE))

cat("\nRows after filter:", nrow(df), "\n")
```

---

# Step 8 — Code factors for downstream modeling

```{r factor-coding}
df <- df %>%
  mutate(
    STAGE    = factor(STAGE, levels = 1:4, labels = c("I", "II", "III", "IV")),
    SUBTYPE  = factor(SUBTYPE, levels = c("LumA", "LumB", "HER2E", "TNBC")),
    # Gene mutations as factors (wild-type = reference)
    across(all_of(final_panel),
           ~ factor(., levels = c(0, 1), labels = c("Wild-type", "Mutated")))
  )

cat("Final class of key variables:\n")
str(df %>% select(OS_MONTHS, OS_EVENT, AGE, STAGE, SUBTYPE,
                  all_of(final_panel)[1:3]))
```

---

# Step 9 — Sanity checks and descriptive overview

```{r sanity-overview}
cat("==== Analytic cohort summary ====\n")
cat("N patients        :", nrow(df), "\n")
cat("Events (deaths)   :", sum(df$OS_EVENT == 1), "\n")
cat("Censored          :", sum(df$OS_EVENT == 0), "\n")
cat("Median follow-up  :", round(median(df$OS_MONTHS, na.rm = TRUE), 1), "months\n\n")

cat("SUBTYPE x STAGE:\n")
print(table(df$SUBTYPE, df$STAGE, useNA = "ifany"))

cat("\nMutation rate per gene (overall):\n")
sapply(df[, final_panel],
       function(x) round(mean(x == "Mutated"), 3)) %>%
  sort(decreasing = TRUE) %>% print()
```

```{r mut-rate-by-subtype}
# Mutation rate per gene x subtype
rate_matrix <- sapply(final_panel, function(g) {
  tapply(df[[g]] == "Mutated", df$SUBTYPE, mean)
})
rate_matrix <- round(rate_matrix * 100, 1)

kable(as.data.frame(rate_matrix),
      caption = "Mutation rate (%) per gene by subtype")
```

---

# Step 10 — Save analytic dataset

```{r save}
saveRDS(df, file.path(dataset_dir, "df_analysis.rds"))

# Also save as CSV for convenience / external inspection (factor levels
# preserved as strings)
write.csv(df %>% mutate(across(where(is.factor), as.character)),
          file.path(dataset_dir, "df_analysis.csv"), row.names = FALSE)

cat("Saved:\n")
cat(" - ", file.path(dataset_dir, "df_analysis.rds"), "\n", sep = "")
cat(" - ", file.path(dataset_dir, "df_analysis.csv"), "\n", sep = "")
```

---

# Notes

- The downstream analysis Rmd (`survival_analysis.Rmd`) should load
  `df_analysis.rds` and the panel via `readRDS("dataset/final_panel.rds")`.
- For the primary-only sensitivity analysis, subset with
  `filter(HAS_PRIMARY == 1)`.
- For the TMB-adjusted sensitivity analysis, include `TMB` as a continuous
  covariate in Cox models (may need log-transform or spline if skewed).
- If the merge produces fewer patients than expected, check the HR / HER2
  missingness — patients with unknown subtype are excluded at Step 7.
