---
title: "Final Analysis — Breast Cancer Gene Mutation & Overall Survival"
subtitle: "Subtype-Stratified Cox Analysis · MSK-CHORD 2024 · BIOSTAT 629"
author: "Yiqiao Zhu"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center")
```

# Setup and Load Data

```{r libraries}
library(dplyr)
library(tidyr)
library(survival)
library(survminer)
library(broom)
library(knitr)
library(ggplot2)
library(purrr)
```

```{r paths-and-load}
proj_dir    <- "/Users/yiqiaozhu/Desktop/BIOS 629"
final_dir   <- file.path(proj_dir, "Final Project")
dataset_dir <- file.path(final_dir, "dataset")
plot_dir    <- file.path(final_dir, "plots")
dir.create(plot_dir, showWarnings = FALSE, recursive = TRUE)

df          <- readRDS(file.path(dataset_dir, "df_analysis.rds"))
final_panel <- readRDS(file.path(dataset_dir, "final_panel.rds"))
gene_terms  <- paste(final_panel, collapse = " + ")

cat("Cohort size :", nrow(df), "\n")
cat("Panel genes :", length(final_panel), "\n")
print(final_panel)
```

---

# Cohort Overview (Table 1 by Subtype)

Stratified descriptive statistics to check baseline balance across the four
clinical subtypes.

```{r table1-overall}
cat("==== Overall cohort ====\n")
cat("N patients :", nrow(df), "\n")
cat("Deaths      :", sum(df$OS_EVENT == 1), "\n")
cat("Censored    :", sum(df$OS_EVENT == 0), "\n")
cat("Median OS (months):", round(median(df$OS_MONTHS), 1), "\n")
cat("Median age        :", round(median(df$AGE), 1), "\n")
```

```{r table1-by-subtype}
t1_overall <- df %>%
  group_by(SUBTYPE) %>%
  summarise(
    N          = n(),
    Deaths     = sum(OS_EVENT == 1),
    Died_pct   = sprintf("%.1f%%", 100 * mean(OS_EVENT == 1)),
    Age_median = round(median(AGE), 1),
    Age_IQR    = paste0(round(quantile(AGE, .25), 0), "-",
                        round(quantile(AGE, .75), 0)),
    Stage_I    = sprintf("%.1f%%", 100 * mean(STAGE == "I")),
    Stage_II   = sprintf("%.1f%%", 100 * mean(STAGE == "II")),
    Stage_III  = sprintf("%.1f%%", 100 * mean(STAGE == "III")),
    Stage_IV   = sprintf("%.1f%%", 100 * mean(STAGE == "IV")),
    .groups    = "drop"
  )
kable(t1_overall, caption = "Table 1a — Baseline characteristics by subtype")
```

```{r mutation-rate-by-subtype}
# Mutation rate (%) per gene × subtype
rate_mat <- sapply(final_panel, function(g) {
  tapply(df[[g]] == "Mutated", df$SUBTYPE, mean) * 100
})
rate_mat <- round(rate_mat, 1)

kable(as.data.frame(rate_mat),
      caption = "Table 1b — Mutation rate (%) per gene by subtype")
```

---

# Kaplan–Meier Curves (Visual Overview)

For each gene, plot 4-panel KM curves (one panel per subtype), comparing
**Mutated vs Wild-type**.

```{r km-loop, fig.width=12, fig.height=10}
make_km_for_gene <- function(gene) {
  plots <- list()
  for (st in levels(df$SUBTYPE)) {
    sub  <- df %>% filter(SUBTYPE == st)
    form <- as.formula(paste0("Surv(OS_MONTHS, OS_EVENT) ~ ", gene))
    fit  <- surv_fit(form, data = sub)   # survminer's surv_fit avoids the symbol-formula bug

    p <- ggsurvplot(
      fit, data = sub,
      palette       = c("#2E9FDF", "#E7B800"),
      legend.labs   = c("Wild-type", "Mutated"),
      legend.title  = gene,
      title         = paste0(st),
      xlab          = "Time (Months)",
      ylab          = "Survival",
      conf.int      = TRUE,
      pval          = TRUE,
      risk.table    = FALSE,
      xlim          = c(0, 120),
      ggtheme       = theme_classic(base_size = 11)
    )
    plots[[st]] <- p   # keep full ggsurvplot object, not $plot
  }
  plots
}

# Generate + arrange for each gene
for (g in final_panel) {
  pl <- make_km_for_gene(g)
  combined <- arrange_ggsurvplots(list(pl$LumA, pl$LumB, pl$HER2E, pl$TNBC),
                                  ncol = 2, nrow = 2,
                                  title = paste0("KM by Subtype — ", g),
                                  print = FALSE)
  # Save (use png device — more reliable for arrange_ggsurvplots gtable output)
  png(file.path(plot_dir, paste0("KM_", g, "_bySubtype.png")),
      width = 12, height = 10, units = "in", res = 200)
  print(combined)
  dev.off()
}

cat("KM plots saved to:", plot_dir, "\n")
```

```{r km-display-tp53, fig.width=12, fig.height=10, echo=FALSE}
# Display one example (TP53) inline
pl <- make_km_for_gene("TP53")
arrange_ggsurvplots(list(pl$LumA, pl$LumB, pl$HER2E, pl$TNBC),
                    ncol = 2, nrow = 2,
                    title = "Example: KM by Subtype — TP53",
                    print = TRUE)
```

---

# Univariate (Crude) Cox Models

Each (gene × subtype) combination fit **without covariates**. Produces a 32-row
table: `length(final_panel) × 4 subtypes`.

```{r univariate-cox}
uni_results <- expand.grid(
  Gene    = final_panel,
  Subtype = levels(df$SUBTYPE),
  KEEP.OUT.ATTRS = FALSE, stringsAsFactors = FALSE
) %>%
  rowwise() %>%
  mutate(
    res = list({
      sub  <- df %>% filter(SUBTYPE == Subtype)
      form <- as.formula(paste0("Surv(OS_MONTHS, OS_EVENT) ~ ", Gene))
      fit  <- coxph(form, data = sub)
      tidy(fit, exponentiate = TRUE, conf.int = TRUE)
    })
  ) %>%
  ungroup() %>%
  unnest(res) %>%
  filter(grepl("Mutated", term)) %>%
  transmute(
    Gene, Subtype,
    HR       = round(estimate, 3),
    CI_low   = round(conf.low,  3),
    CI_high  = round(conf.high, 3),
    p.value  = signif(p.value,  4)
  )

kable(uni_results, caption = "Univariate Cox — crude HR per gene × subtype")
```

```{r univariate-wide}
# Wide format: genes as rows, subtypes as columns (HR only)
uni_wide <- uni_results %>%
  mutate(HR_fmt = sprintf("%.2f (%.2f–%.2f)%s",
                          HR, CI_low, CI_high,
                          ifelse(p.value < 0.05, "*", ""))) %>%
  select(Gene, Subtype, HR_fmt) %>%
  pivot_wider(names_from = Subtype, values_from = HR_fmt)

kable(uni_wide,
      caption = "Univariate Cox HR (95% CI) per gene × subtype ( * = p < 0.05 )")
```

# Visulization for Unvariate Cox model
```{r}
hm_crude <- uni_results %>%
  mutate(
    logHR = log(HR),
    sig   = ifelse(p.value < 0.05, "*", ""),
    HR_label = ifelse(HR == 0 | is.infinite(CI_high),
                      "N/E",
                      sprintf("%.2f%s", HR, sig)),
    logHR = ifelse(HR_label == "N/E", NA, logHR),
    Gene    = factor(Gene, levels = rev(final_panel)),
    Subtype = factor(Subtype, levels = levels(df$SUBTYPE))
  )

p_crude <- ggplot(hm_crude, aes(x = Subtype, y = Gene, fill = logHR)) +
  geom_tile(color = "white") +
  geom_text(aes(label = HR_label), size = 3.5) +
  scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                       midpoint = 0, na.value = "gray70", name = "log(HR)") +
  labs(title = "Crude HR (Unadjusted)",
       subtitle = "Univariate Cox per gene × subtype (* = p < 0.05, N/E = not estimable)",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank())

print(p_crude)

ggsave(file.path(plot_dir, "HR_heatmap_crude.png"),
       p_crude, width = 8, height = 6, dpi = 300)
```


---

---

# Main Analysis — Subtype-Stratified Multivariable Cox

Main result. For each subtype separately, fit:
**Surv(OS) ~ all 8 genes + AGE + STAGE**. Produces subtype-specific adjusted
HRs for every gene.

```{r stratified-cox-loop}
fit_strat <- function(st) {
  sub  <- df %>% filter(SUBTYPE == st)
  form <- as.formula(paste0("Surv(OS_MONTHS, OS_EVENT) ~ ",
                            gene_terms, " + AGE + STAGE"))
  fit  <- coxph(form, data = sub)
  list(fit = fit, n = nrow(sub), events = sum(sub$OS_EVENT == 1),
       tidy_tbl = tidy(fit, exponentiate = TRUE, conf.int = TRUE))
}

strat_list <- lapply(levels(df$SUBTYPE), fit_strat)
names(strat_list) <- levels(df$SUBTYPE)

# Cohort size / events per subtype
sizes <- data.frame(
  Subtype = names(strat_list),
  N       = sapply(strat_list, function(x) x$n),
  Events  = sapply(strat_list, function(x) x$events)
)
kable(sizes, caption = "Subtype cohort sizes and event counts")
```

```{r stratified-cox-summary}
# Long format: all gene HRs across subtypes
strat_results <- map2_dfr(strat_list, names(strat_list), function(x, st) {
  x$tidy_tbl %>%
    filter(grepl("Mutated", term)) %>%
    mutate(
      Subtype = st,
      Gene    = gsub("Mutated$", "", term),
      HR      = round(estimate, 3),
      CI_low  = round(conf.low,  3),
      CI_high = round(conf.high, 3),
      p.value = signif(p.value,  4)
    ) %>%
    select(Gene, Subtype, HR, CI_low, CI_high, p.value)
})

kable(strat_results,
      caption = "Subtype-stratified multivariable Cox — adjusted HR per gene × subtype")
```

```{r stratified-wide}
strat_wide <- strat_results %>%
  mutate(HR_fmt = sprintf("%.2f (%.2f–%.2f)%s",
                          HR, CI_low, CI_high,
                          ifelse(p.value < 0.05, "*", ""))) %>%
  select(Gene, Subtype, HR_fmt) %>%
  pivot_wider(names_from = Subtype, values_from = HR_fmt)

kable(strat_wide,
      caption = "Adjusted HR (95% CI) per gene × subtype ( * = p < 0.05 )")
```

---

# Visualization of Main Results

## Heatmap — log(HR) across genes × subtypes

```{r heatmap, fig.width=8, fig.height=6}
hm_adj <- strat_results %>%
  mutate(
    logHR = log(HR),
    sig   = ifelse(p.value < 0.05, "*", ""),
    HR_label = ifelse(HR == 0 | is.infinite(CI_high),
                      "N/E",
                      sprintf("%.2f%s", HR, sig)),
    logHR = ifelse(HR_label == "N/E", NA, logHR),
    Gene    = factor(Gene, levels = rev(final_panel)),
    Subtype = factor(Subtype, levels = levels(df$SUBTYPE))
  )

p_adj <- ggplot(hm_adj, aes(x = Subtype, y = Gene, fill = logHR)) +
  geom_tile(color = "white") +
  geom_text(aes(label = HR_label), size = 3.5) +
  scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                       midpoint = 0, na.value = "gray70", name = "log(HR)") +
  labs(title = "Adjusted HR — Gene × Subtype",
       subtitle = "Multivariable Cox (adjusted for age + stage); * = p < 0.05, N/E = not estimable",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank())

print(p_adj)

ggsave(file.path(plot_dir, "HR_heatmap_adjusted.png"),
       p_adj, width = 8, height = 6, dpi = 300)
```

## Forest plot — per-subtype

```{r forest-per-subtype, fig.width=11, fig.height=9}
forest_df <- strat_results %>%
  filter(HR > 0, !is.infinite(CI_high), !is.na(HR)) %>%   # ← 新增：过滤 N/E
  mutate(Gene = factor(Gene, levels = rev(final_panel)),
         sig  = ifelse(p.value < 0.05, "Significant", "Not significant"))

ggplot(forest_df, aes(x = HR, y = Gene, color = sig)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray50") +
  geom_errorbarh(aes(xmin = CI_low, xmax = CI_high),
                 height = 0.25, linewidth = 0.8) +
  geom_point(size = 3) +
  scale_color_manual(values = c("Significant"     = "#C0392B",
                                "Not significant" = "#7F8C8D"),
                     name = NULL) +
  facet_wrap(~ Subtype, scales = "free_x", ncol = 2) +
  scale_x_log10() +
  labs(title    = "Forest Plot — Adjusted HR by Subtype",
       subtitle = "Multivariable Cox per subtype, adjusted for age + stage (N/E combinations excluded)",
       x = "Hazard Ratio (log scale)", y = NULL) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom",
        strip.background = element_rect(fill = "#003366", color = NA),
        strip.text       = element_text(color = "white", face = "bold"))

ggsave(file.path(plot_dir, "Forest_bySubtype.png"),
       width = 11, height = 9, dpi = 300)
```

---

# Summary of Main Findings

```{r summary-findings}
# Significant gene × subtype combinations (p < 0.05)
sig_hits <- strat_results %>%
  filter(p.value < 0.05) %>%
  arrange(Subtype, p.value)

kable(sig_hits,
      caption = paste0("Significant gene × subtype combinations (p < 0.05) — ",
                       nrow(sig_hits), " hits from ",
                       nrow(strat_results), " tests"))
```

```{r top-hits}
# Top 5 by smallest p-value
top_hits <- strat_results %>% arrange(p.value) %>% head(5)
kable(top_hits,
      caption = "Top 5 gene × subtype combinations by p-value")
```

---

# Multiple Testing Correction (BH)

The 32 gene × subtype adjusted HRs generate 32 simultaneous tests; apply
Benjamini–Hochberg correction to control FDR.

```{r bh-correction}
strat_results <- strat_results %>%
  mutate(q_BH = signif(p.adjust(p.value, method = "BH"), 4),
         sig_raw = ifelse(p.value < 0.05, "*", ""),
         sig_BH  = ifelse(q_BH    < 0.05, "*", ""))

kable(strat_results %>% arrange(q_BH),
      caption = "Subtype-stratified HR with BH-adjusted q-values (32 tests)")
```

```{r bh-summary}
n_raw <- sum(strat_results$p.value < 0.05)
n_bh  <- sum(strat_results$q_BH    < 0.05)
cat("Significant (raw p < 0.05) :", n_raw, "/", nrow(strat_results), "\n")
cat("Significant (BH q < 0.05)  :", n_bh,  "/", nrow(strat_results), "\n\n")

kable(strat_results %>% filter(q_BH < 0.05) %>% arrange(q_BH),
      caption = "Gene × Subtype combinations surviving BH (q < 0.05)")
```

---

---

# Proportional Hazards Assumption Check

Schoenfeld residual test (`cox.zph`) for each subtype-stratified Cox model.
Per-variable p < 0.05 indicates PH violation for that variable.

```{r ph-check}
# Focus: PH check only for the 4 BH-significant hits
hits <- data.frame(
  Gene    = c("ESR1", "TP53", "CDH1", "TP53"),
  Subtype = c("LumA", "LumA", "LumB", "TNBC")
)

ph_hits <- purrr::pmap_dfr(hits, function(Gene, Subtype) {
  fit <- strat_list[[Subtype]]$fit
  zph <- cox.zph(fit)
  row <- as.data.frame(zph$table) %>%
    tibble::rownames_to_column("variable") %>%
    filter(variable == Gene)
  data.frame(
    Gene      = Gene,
    Subtype   = Subtype,
    chisq     = round(row$chisq, 3),
    p_value   = signif(row$p, 4),
    PH_status = ifelse(row$p < 0.05, "Violated", "Not violated")
  )
})

kable(ph_hits,
      caption = "PH assumption check for the 4 BH-significant gene × subtype hits")
```

---


---

# Sensitivity Analysis  — TMB Adjustment

Adjust for tumor mutational burden (log-transformed) to check whether gene
effects are independent of overall mutation load.

```{r sens-tmb}
df_tmb <- df %>%
  filter(!is.na(TMB)) %>%
  mutate(log_TMB = log(TMB + 1))   # add 1 to avoid log(0)

cat("TMB-adjustable cohort size:", nrow(df_tmb),
    "(", round(nrow(df_tmb) / nrow(df) * 100, 1), "% of full)\n\n")

strat_tmb <- lapply(levels(df_tmb$SUBTYPE), function(st) {
  sub  <- df_tmb %>% filter(SUBTYPE == st)
  form <- as.formula(paste0("Surv(OS_MONTHS, OS_EVENT) ~ ",
                            gene_terms, " + AGE + STAGE + log_TMB"))
  fit  <- coxph(form, data = sub)
  tidy(fit, exponentiate = TRUE, conf.int = TRUE) %>%
    filter(grepl("Mutated", term)) %>%
    mutate(Subtype = st,
           Gene    = gsub("Mutated$", "", term),
           HR      = round(estimate, 3),
           p.value = signif(p.value, 4)) %>%
    select(Gene, Subtype, HR, p.value)
}) %>% bind_rows()

compare_tmb <- strat_results %>%
  select(Gene, Subtype, HR_main = HR, p_main = p.value) %>%
  left_join(strat_tmb %>%
              rename(HR_tmb = HR, p_tmb = p.value),
            by = c("Gene", "Subtype"))

kable(compare_tmb,
      caption = "Main analysis vs TMB-adjusted: HR comparison")
```

# Visulization for TMP adjustment
```{r}
# Define BH hits (in case not defined upstream)
bh_hits <- c("ESR1_LumA", "TP53_LumA", "CDH1_LumB", "TP53_TNBC")

# 1. Stratified Cox with TMB adjustment (keep CI)
strat_tmb <- lapply(levels(df_tmb$SUBTYPE), function(st) {
  sub  <- df_tmb %>% filter(SUBTYPE == st)
  form <- as.formula(paste0("Surv(OS_MONTHS, OS_EVENT) ~ ",
                            gene_terms, " + AGE + STAGE + log_TMB"))
  fit  <- suppressWarnings(coxph(form, data = sub))
  tidy(fit, exponentiate = TRUE, conf.int = TRUE) %>%
    filter(grepl("Mutated", term)) %>%
    mutate(Subtype = st,
           Gene    = gsub("Mutated$", "", term),
           HR      = round(estimate, 3),
           CI_low  = round(conf.low, 3),
           CI_high = round(conf.high, 3),
           p.value = signif(p.value, 4)) %>%
    select(Gene, Subtype, HR, CI_low, CI_high, p.value)
}) %>% bind_rows()

compare_tmb <- strat_results %>%
  select(Gene, Subtype, HR_main = HR,
         CIlo_main = CI_low, CIhi_main = CI_high, p_main = p.value) %>%
  left_join(strat_tmb %>%
              select(Gene, Subtype, HR_tmb = HR,
                     CIlo_tmb = CI_low, CIhi_tmb = CI_high, p_tmb = p.value),
            by = c("Gene", "Subtype"))

# 2. Dumbbell plot with CI, 4 BH hits
plot_df_tmb <- compare_tmb %>%
  mutate(id = paste(Gene, Subtype, sep = "_")) %>%
  filter(id %in% bh_hits) %>%
  pivot_longer(
    cols = c(HR_main, HR_tmb, CIlo_main, CIlo_tmb, CIhi_main, CIhi_tmb),
    names_to = c(".value", "Analysis"),
    names_pattern = "(HR|CIlo|CIhi)_(main|tmb)"
  ) %>%
  mutate(Analysis = factor(Analysis,
                           levels = c("main", "tmb"),
                           labels = c("Main", "TMB-adjusted")),
         label = paste(Gene, Subtype, sep = " × "))

ggplot(plot_df_tmb, aes(x = HR, y = label, color = Analysis)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray50") +
  geom_errorbarh(aes(xmin = CIlo, xmax = CIhi),
                 height = 0.2, linewidth = 0.7,
                 position = position_dodge(width = 0.4)) +
  geom_point(size = 3.5, position = position_dodge(width = 0.4)) +
  scale_color_manual(values = c("Main" = "#003366", "TMB-adjusted" = "#16A085")) +
  scale_x_log10(
    breaks = c(1, 1.5, 2, 2.5, 3, 3.5, 4, 5),
    labels = c("1", "1.5", "2", "2.5", "3", "3.5", "4", "5")
  ) +
  labs(title = "Sensitivity — BH-significant results: Main vs TMB-adjusted",
       subtitle = "Error bars = 95% CI",
       x = "Hazard Ratio (log scale)", y = NULL) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")
```

```{r}
ggsave(file.path(plot_dir, "Sens_tmb_dumbbell_CI.png"),
       width = 8, height = 6, dpi = 300)
```
---

---

# Restricted Mean Survival Time (RMST)

RMST difference at τ = 60 months (5 years) as an alternative effect measure.
Does not rely on the PH assumption; results are interpretable in absolute
time units.

```{r rmst-compute}
library(survRM2)
tau_months <- 60

hits <- data.frame(
  Gene    = c("ESR1", "TP53", "CDH1", "TP53"),
  Subtype = c("LumA", "LumA", "LumB", "TNBC")
)

rmst_bh <- purrr::pmap_dfr(hits, function(Gene, Subtype) {
  sub <- df %>% filter(SUBTYPE == Subtype)
  arm <- ifelse(sub[[Gene]] == "Mutated", 1L, 0L)
  r <- rmst2(time = sub$OS_MONTHS, status = sub$OS_EVENT,
             arm = arm, tau = tau_months)
  d <- r$unadjusted.result["RMST (arm=1)-(arm=0)", ]
  data.frame(
    Gene, Subtype,
    RMST_diff = round(d["Est."], 2),
    CI_low    = round(d["lower .95"], 2),
    CI_high   = round(d["upper .95"], 2),
    p_value   = signif(d["p"], 4)
  )
})

# Join with Cox HR for side-by-side
rmst_hr_compare <- strat_results %>%
  semi_join(hits, by = c("Gene", "Subtype")) %>%
  select(Gene, Subtype, HR, HR_p = p.value) %>%
  left_join(rmst_bh, by = c("Gene", "Subtype")) %>%
  mutate(Concordant = ifelse((HR > 1 & RMST_diff < 0) |
                             (HR < 1 & RMST_diff > 0), "✓", "✗"))

kable(rmst_hr_compare,
      caption = paste0("RMST validation of BH-significant hits (τ = ",
                       tau_months, " months)"))
```

---

# Summary & Limitations

```{r final-summary}
final_summary <- strat_results %>%
  semi_join(hits, by = c("Gene", "Subtype")) %>%
  select(Gene, Subtype,
         HR, CI_low, CI_high, p_raw = p.value, q_BH) %>%
  left_join(ph_hits %>% select(Gene, Subtype, PH_status),
            by = c("Gene", "Subtype")) %>%
  left_join(rmst_bh %>% select(Gene, Subtype,
                                RMST_months = RMST_diff,
                                RMST_p = p_value),
            by = c("Gene", "Subtype")) %>%
  left_join(compare_tmb %>% select(Gene, Subtype, HR_tmb),
            by = c("Gene", "Subtype")) %>%
  arrange(q_BH)

kable(final_summary,
      caption = "Integrated summary of 4 BH-significant gene × subtype hits with all validations")
```


