---
title: "Cancer Driver Gene Mutations and Subtype-Specific Effects on Overall Survival in Breast Cancer"
subtitle: "Integrated Analysis Pipeline · MSK-CHORD 2024 · BIOSTAT 629 Final Project"
author: |
  Yiqiao Zhu  
  [GitHub Repository](https://github.com/Ricardo-z123/BIOSTAT629-Final-Project) · [Website](https://ricardo-z123.github.io/BIOSTAT629-Final-Project-website/)
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    code_folding: show
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center")
set.seed(42)   # fixed seed for reproducibility
```

# Setup

## Libraries

```{r}
required_pkgs <- c("dplyr", "tidyr", "readr", "knitr", "purrr", "ggplot2",
                   "survival", "survminer", "broom", "survRM2", "tibble")

install.packages(required_pkgs, repos = "https://cloud.r-project.org")

```

```{r libraries}
library(dplyr)
library(tidyr)
library(readr)
library(knitr)
library(purrr)
library(ggplot2)
library(survival)
library(survminer)
library(broom)
library(survRM2)
library(tibble)
```

## Paths

```{r paths}
# ---- Portable project paths (relative to this Rmd) ------------------------
# When knitting an Rmd, R's working directory is the Rmd's own folder by
# default, so all relative paths anchor to the location of this file.
dataset_dir <- "dataset"
plot_dir    <- "plots"
dir.create(dataset_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(plot_dir,    showWarnings = FALSE, recursive = TRUE)

# ---- COSMIC (local; auto-detect across common locations) ------------------
cosmic_filename <- "Cosmic_CancerGeneCensus_v103_GRCh38.tsv"

cosmic_candidates <- c(
  cosmic_filename,                                                       # same dir as Rmd
  file.path("Cosmic_CancerGeneCensus_Tsv_v103_GRCh38", cosmic_filename), # subfolder
  file.path("cosmic", cosmic_filename),                                  # generic subfolder
  file.path("data",   cosmic_filename)                                   # generic subfolder
)

cosmic_path <- cosmic_candidates[file.exists(cosmic_candidates)][1]

if (is.na(cosmic_path)) {
  stop("COSMIC file not found. Please place '", cosmic_filename,
       "' in the same folder as this Rmd file.")
}
cat("Using COSMIC file:", cosmic_path, "\n")

# ---- MSK-CHORD 2024 (remote, cBioPortal datahub) --------------------------
msk_base <- paste0(
  "https://media.githubusercontent.com/media/cBioPortal/datahub/",
  "refs/heads/master/public/msk_chord_2024/"
)

patient_url <- paste0(msk_base, "data_clinical_patient.txt")
sample_url  <- paste0(msk_base, "data_clinical_sample.txt")
mut_url     <- paste0(msk_base, "data_mutations.txt")
```

## Load raw data

```{r load-raw}
# COSMIC Cancer Gene Census (local)
cgc <- read.delim(cosmic_path, sep = "\t", header = TRUE,
                  stringsAsFactors = FALSE, quote = "")

# MSK-CHORD 2024 (remote)
pat  <- read.delim(patient_url, comment.char = "#",
                   stringsAsFactors = TRUE, check.names = FALSE)
samp <- read.delim(sample_url,  comment.char = "#",
                   stringsAsFactors = TRUE, check.names = FALSE)
mut  <- read.delim(mut_url,     comment.char = "#",
                   stringsAsFactors = TRUE, check.names = FALSE)

cat("CGC genes      :", nrow(cgc),  "\n")
cat("Patient rows   :", nrow(pat),  "\n")
cat("Sample rows    :", nrow(samp), "\n")
cat("Mutation rows  :", nrow(mut),  "\n")
```

------------------------------------------------------------------------

# Part 1 — Gene Panel Selection

Construct an evidence-based breast cancer driver gene panel by
intersecting:

1.  **COSMIC Cancer Gene Census v103** — externally curated cancer
    driver genes whose `TUMOUR_TYPES_SOMATIC` field includes "breast",
    restricted to Tier-1 evidence.
2.  **MSK-CHORD 2024 breast cohort** — genes with patient-level mutation
    rate $\geq 5\%$ (ensures power for subtype-stratified analyses).

The panel is the **intersection** of the two lists.

## Filter COSMIC for breast cancer drivers

```{r cosmic-breast}
breast_drivers <- cgc %>%
  filter(grepl("breast", TUMOUR_TYPES_SOMATIC, ignore.case = TRUE)) %>%
  select(GENE_SYMBOL, TIER, ROLE_IN_CANCER, TUMOUR_TYPES_SOMATIC)

cat("Breast-related drivers in CGC:", nrow(breast_drivers), "\n")

# Tier-1 only (strong evidence)
breast_drivers_tier1 <- breast_drivers %>% filter(TIER == "1")
cat("Breast Tier-1 drivers        :", nrow(breast_drivers_tier1), "\n")
```

## MSK-CHORD breast cohort: patient-level mutation frequencies

```{r msk-freq}
bc_samp <- samp %>% filter(CANCER_TYPE == "Breast Cancer")
bc_ids  <- unique(bc_samp$PATIENT_ID)

cat("BC samples :", nrow(bc_samp),     "\n")
cat("BC patients:", length(bc_ids),    "\n")

bc_mut <- mut %>%
  filter(Tumor_Sample_Barcode %in% bc_samp$SAMPLE_ID) %>%
  inner_join(bc_samp %>% select(SAMPLE_ID, PATIENT_ID),
             by = c("Tumor_Sample_Barcode" = "SAMPLE_ID"))

gene_freq <- bc_mut %>%
  distinct(PATIENT_ID, Hugo_Symbol) %>%
  count(Hugo_Symbol, name = "n_mut") %>%
  mutate(rate_pct = round(n_mut / length(bc_ids) * 100, 2)) %>%
  arrange(desc(n_mut))

freq_threshold <- 5  # in percent
msk_freq_panel <- gene_freq %>%
  filter(rate_pct >= freq_threshold) %>%
  pull(Hugo_Symbol)

cat("Threshold: mutation rate >=", freq_threshold, "%\n")
cat("MSK BC genes passing threshold:", length(msk_freq_panel), "\n")
print(msk_freq_panel)
```

## Intersection → final 8-gene panel

```{r intersection}
final_panel <- intersect(breast_drivers_tier1$GENE_SYMBOL, msk_freq_panel)

cat("==== FINAL PANEL (COSMIC Tier-1 breast drivers ∩ MSK freq >=",
    freq_threshold, "%) ====\n")
cat("N =", length(final_panel), "genes\n")
print(final_panel)
```

```{r panel-summary}
panel_summary <- gene_freq %>%
  filter(Hugo_Symbol %in% final_panel) %>%
  left_join(
    breast_drivers %>% select(GENE_SYMBOL, TIER, ROLE_IN_CANCER),
    by = c("Hugo_Symbol" = "GENE_SYMBOL")
  ) %>%
  arrange(desc(rate_pct))

kable(panel_summary,
      caption = "Final gene panel with mutation frequency and CGC role")

# Persist for reproducibility
saveRDS(final_panel, file.path(dataset_dir, "final_panel.rds"))
write.csv(panel_summary,
          file.path(dataset_dir, "final_panel_summary.csv"),
          row.names = FALSE)
```

------------------------------------------------------------------------

# Part 2 — Analytic Dataset Construction

Build the analytic dataset (`df_analysis`) used for all downstream
survival modeling, containing:

-   **Outcome**: `OS_MONTHS`, `OS_EVENT`
-   **Primary exposures**: 8 binary patient-level mutation indicators
-   **Effect modifier**: `SUBTYPE` (LumA / LumB / HER2E / TNBC)
-   **Confounders**: `AGE`, `STAGE` (I–IV)
-   **Sensitivity-analysis variables**: `TMB`

## Patient-level mutation indicators (panel genes)

A patient is mutated (1) for a gene if **any** of their tumor samples
harbors a mutation in that gene; wild-type (0) otherwise.

```{r mutation-indicators}
bc_mut_panel <- mut %>%
  filter(Tumor_Sample_Barcode %in% bc_samp$SAMPLE_ID,
         Hugo_Symbol %in% final_panel) %>%
  inner_join(bc_samp %>% select(SAMPLE_ID, PATIENT_ID),
             by = c("Tumor_Sample_Barcode" = "SAMPLE_ID"))

mut_wide <- bc_mut_panel %>%
  distinct(PATIENT_ID, Hugo_Symbol) %>%
  mutate(x = 1L) %>%
  pivot_wider(names_from = Hugo_Symbol, values_from = x, values_fill = 0L)

# Ensure every panel gene has a column (even if 0 mutations)
for (g in final_panel) {
  if (!g %in% names(mut_wide)) mut_wide[[g]] <- 0L
}
mut_wide <- mut_wide[, c("PATIENT_ID", final_panel)]

cat("Patients with >=1 mutation in panel genes:", nrow(mut_wide), "\n")
```

## Aggregate sample-level attributes to patient level

Highest STAGE across samples (worst-case clinical interpretation); max
TMB across samples

```{r sample-aggregation}
recode_stage <- function(x) {
  dplyr::case_when(
    x %in% c("1", "1A", "1B")  ~ 1L,
    x %in% c("2A", "2B")       ~ 2L,
    x %in% c("3A", "3B", "3C") ~ 3L,
    x == "4"                   ~ 4L,
    TRUE                       ~ NA_integer_
  )
}

bc_samp2 <- bc_samp %>%
  mutate(STAGE_INT = recode_stage(as.character(CLINICAL_GROUP)),
         TMB_NUM   = suppressWarnings(as.numeric(as.character(TMB_NONSYNONYMOUS))))

sample_agg <- bc_samp2 %>%
  group_by(PATIENT_ID) %>%
  summarise(
    STAGE       = suppressWarnings(max(STAGE_INT, na.rm = TRUE)),
    TMB         = suppressWarnings(max(TMB_NUM,   na.rm = TRUE)),
    HAS_PRIMARY = as.integer(any(SAMPLE_TYPE == "Primary", na.rm = TRUE)),
    N_SAMPLES   = n(),
    .groups     = "drop"
  ) %>%
  mutate(
    STAGE = ifelse(is.infinite(STAGE), NA_integer_, STAGE),
    TMB   = ifelse(is.infinite(TMB),   NA_real_,    TMB)
  )
```

## Clean patient-level clinical data + derive SUBTYPE

```{r clinical-clean}
pat_bc <- pat %>%
  filter(PATIENT_ID %in% bc_ids) %>%
  select(PATIENT_ID, OS_MONTHS, OS_STATUS,
         CURRENT_AGE_DEID, RACE, HR, HER2) %>%
  rename(AGE = CURRENT_AGE_DEID) %>%
  mutate(
    OS_MONTHS = suppressWarnings(as.numeric(as.character(OS_MONTHS))),
    AGE       = suppressWarnings(as.numeric(as.character(AGE))),
    OS_EVENT = dplyr::case_when(
      OS_STATUS == "1:DECEASED" ~ 1L,
      OS_STATUS == "0:LIVING"   ~ 0L,
      TRUE                      ~ NA_integer_
    ),
    # SUBTYPE from HR x HER2
    SUBTYPE = dplyr::case_when(
      HR == "Yes" & HER2 == "No"  ~ "LumA",
      HR == "Yes" & HER2 == "Yes" ~ "LumB",
      HR == "No"  & HER2 == "Yes" ~ "HER2E",
      HR == "No"  & HER2 == "No"  ~ "TNBC",
      TRUE                         ~ NA_character_
    )
  )

cat("HR distribution:\n");      print(table(pat_bc$HR,      useNA = "ifany"))
cat("\nHER2 distribution:\n");  print(table(pat_bc$HER2,    useNA = "ifany"))
cat("\nSUBTYPE distribution:\n"); print(table(pat_bc$SUBTYPE, useNA = "ifany"))
```

## Merge components, filter, and code factors

```{r merge-filter-factor}
df <- pat_bc %>%
  left_join(mut_wide,   by = "PATIENT_ID") %>%
  left_join(sample_agg, by = "PATIENT_ID") %>%
  # patients without any mutation record -> wild-type for all panel genes
  mutate(across(all_of(final_panel),
                ~ ifelse(is.na(.), 0L, as.integer(.))))

cat("Rows after merge:", nrow(df), "\n")

cat("\nMissingness BEFORE filter:\n")
sapply(df %>% select(OS_MONTHS, OS_EVENT, AGE, STAGE, SUBTYPE),
       function(x) sum(is.na(x))) %>% print()

df <- df %>%
  filter(!is.na(OS_MONTHS), !is.na(OS_EVENT),
         !is.na(AGE),       !is.na(STAGE),
         !is.na(SUBTYPE))

cat("\nRows after filter:", nrow(df), "\n")

# Factor coding for downstream modeling
df <- df %>%
  mutate(
    STAGE   = factor(STAGE, levels = 1:4, labels = c("I", "II", "III", "IV")),
    SUBTYPE = factor(SUBTYPE, levels = c("LumA", "LumB", "HER2E", "TNBC")),
    across(all_of(final_panel),
           ~ factor(., levels = c(0, 1), labels = c("Wild-type", "Mutated")))
  )

# Persist analytic dataset
saveRDS(df, file.path(dataset_dir, "df_analysis.rds"))
write.csv(df %>% mutate(across(where(is.factor), as.character)),
          file.path(dataset_dir, "df_analysis.csv"), row.names = FALSE)
```

## Cohort overview

```{r cohort-overview}
cat("==== Analytic cohort summary ====\n")
cat("N patients        :", nrow(df), "\n")
cat("Events (deaths)   :", sum(df$OS_EVENT == 1), "\n")
cat("Censored          :", sum(df$OS_EVENT == 0), "\n")
cat("Median follow-up  :", round(median(df$OS_MONTHS, na.rm = TRUE), 1), "months\n\n")

cat("SUBTYPE distribution:\n")
print(table(df$SUBTYPE, useNA = "ifany"))

cat("\nMutation rate per gene (overall):\n")
sapply(df[, final_panel],
       function(x) round(mean(x == "Mutated"), 3)) %>%
  sort(decreasing = TRUE) %>% print()
```

```{r mut-rate-by-subtype}
rate_matrix <- sapply(final_panel, function(g) {
  tapply(df[[g]] == "Mutated", df$SUBTYPE, mean)
})
rate_matrix <- round(rate_matrix * 100, 1)
kable(as.data.frame(rate_matrix),
      caption = "Mutation rate (%) per gene by subtype")
```

------------------------------------------------------------------------

# Part 3 — Survival Analysis

## Cohort Overview (Table 1 by subtype)

```{r table1-by-subtype}
gene_terms <- paste(final_panel, collapse = " + ")

t1 <- df %>%
  group_by(SUBTYPE) %>%
  summarise(
    N          = n(),
    Deaths     = sum(OS_EVENT == 1),
    Died_pct   = sprintf("%.1f%%", 100 * mean(OS_EVENT == 1)),
    Age_median = round(median(AGE), 1),
    Age_IQR    = paste0(round(quantile(AGE, .25), 0), "-",
                        round(quantile(AGE, .75), 0)),
    Stage_I    = sprintf("%.1f%%", 100 * mean(STAGE == "I")),
    Stage_II   = sprintf("%.1f%%", 100 * mean(STAGE == "II")),
    Stage_III  = sprintf("%.1f%%", 100 * mean(STAGE == "III")),
    Stage_IV   = sprintf("%.1f%%", 100 * mean(STAGE == "IV")),
    .groups    = "drop"
  )
kable(t1, caption = "Table 1 — Baseline characteristics by subtype")
```

## Kaplan–Meier curves (4-panel by subtype, per gene)

```{r km-loop, fig.width=12, fig.height=10}
make_km_for_gene <- function(gene) {
  plots <- list()
  for (st in levels(df$SUBTYPE)) {
    sub  <- df %>% filter(SUBTYPE == st)
    form <- as.formula(paste0("Surv(OS_MONTHS, OS_EVENT) ~ ", gene))
    fit  <- surv_fit(form, data = sub)

    p <- ggsurvplot(
      fit, data = sub,
      palette       = c("#2E9FDF", "#E7B800"),
      legend.labs   = c("Wild-type", "Mutated"),
      legend.title  = gene,
      title         = st,
      xlab          = "Time (Months)",
      ylab          = "Survival",
      conf.int      = TRUE,
      pval          = TRUE,
      risk.table    = FALSE,
      xlim          = c(0, 120),
      ggtheme       = theme_classic(base_size = 11)
    )
    plots[[st]] <- p
  }
  plots
}

for (g in final_panel) {
  pl <- make_km_for_gene(g)
  combined <- arrange_ggsurvplots(list(pl$LumA, pl$LumB, pl$HER2E, pl$TNBC),
                                  ncol = 2, nrow = 2,
                                  title = paste0("KM by Subtype — ", g),
                                  print = FALSE)
  png(file.path(plot_dir, paste0("KM_", g, "_bySubtype.png")),
      width = 12, height = 10, units = "in", res = 200)
  print(combined)
  dev.off()
}

cat("KM plots saved to:", plot_dir, "\n")
```

```{r km-display-tp53, fig.width=12, fig.height=10, echo=FALSE}
pl <- make_km_for_gene("TP53")
arrange_ggsurvplots(list(pl$LumA, pl$LumB, pl$HER2E, pl$TNBC),
                    ncol = 2, nrow = 2,
                    title = "KM by Subtype — TP53",
                    print = TRUE)
```

## Univariate (crude) Cox models — gene × subtype

```{r univariate-cox}
uni_results <- expand.grid(
  Gene    = final_panel,
  Subtype = levels(df$SUBTYPE),
  KEEP.OUT.ATTRS = FALSE, stringsAsFactors = FALSE
) %>%
  rowwise() %>%
  mutate(
    res = list({
      sub  <- df %>% filter(SUBTYPE == Subtype)
      form <- as.formula(paste0("Surv(OS_MONTHS, OS_EVENT) ~ ", Gene))
      fit  <- coxph(form, data = sub)
      tidy(fit, exponentiate = TRUE, conf.int = TRUE)
    })
  ) %>%
  ungroup() %>%
  unnest(res) %>%
  filter(grepl("Mutated", term)) %>%
  transmute(
    Gene, Subtype,
    HR       = round(estimate, 3),
    CI_low   = round(conf.low,  3),
    CI_high  = round(conf.high, 3),
    p.value  = signif(p.value,  4)
  )

kable(uni_results, caption = "Univariate Cox — crude HR per gene × subtype")
```

```{r heatmap-crude, fig.width=8, fig.height=6}
hm_crude <- uni_results %>%
  mutate(
    logHR = log(HR),
    sig   = ifelse(p.value < 0.05, "*", ""),
    HR_label = ifelse(HR == 0 | is.infinite(CI_high),
                      "N/E", sprintf("%.2f%s", HR, sig)),
    logHR = ifelse(HR_label == "N/E", NA, logHR),
    Gene    = factor(Gene, levels = rev(final_panel)),
    Subtype = factor(Subtype, levels = levels(df$SUBTYPE))
  )

p_crude <- ggplot(hm_crude, aes(x = Subtype, y = Gene, fill = logHR)) +
  geom_tile(color = "white") +
  geom_text(aes(label = HR_label), size = 3.5) +
  scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                       midpoint = 0, na.value = "gray70", name = "log(HR)") +
  labs(title = "Crude HR (Unadjusted)",
       subtitle = "Univariate Cox per gene × subtype (* = p < 0.05, N/E = not estimable)",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank())

print(p_crude)
ggsave(file.path(plot_dir, "HR_heatmap_crude.png"),
       p_crude, width = 8, height = 6, dpi = 300)
```

## Subtype-stratified multivariable Cox (main analysis)

```{r stratified-cox}
fit_strat <- function(st) {
  sub  <- df %>% filter(SUBTYPE == st)
  form <- as.formula(paste0("Surv(OS_MONTHS, OS_EVENT) ~ ",
                            gene_terms, " + AGE + STAGE"))
  fit  <- coxph(form, data = sub)
  list(fit = fit, n = nrow(sub), events = sum(sub$OS_EVENT == 1),
       tidy_tbl = tidy(fit, exponentiate = TRUE, conf.int = TRUE))
}

strat_list <- lapply(levels(df$SUBTYPE), fit_strat)
names(strat_list) <- levels(df$SUBTYPE)

sizes <- data.frame(
  Subtype = names(strat_list),
  N       = sapply(strat_list, function(x) x$n),
  Events  = sapply(strat_list, function(x) x$events)
)
kable(sizes, caption = "Subtype cohort sizes and event counts")

strat_results <- map2_dfr(strat_list, names(strat_list), function(x, st) {
  x$tidy_tbl %>%
    filter(grepl("Mutated", term)) %>%
    mutate(
      Subtype = st,
      Gene    = gsub("Mutated$", "", term),
      HR      = round(estimate, 3),
      CI_low  = round(conf.low,  3),
      CI_high = round(conf.high, 3),
      p.value = signif(p.value,  4)
    ) %>%
    select(Gene, Subtype, HR, CI_low, CI_high, p.value)
})

kable(strat_results,
      caption = "Subtype-stratified multivariable Cox — adjusted HR per gene × subtype")
```

```{r heatmap-adjusted, fig.width=8, fig.height=6}
hm_adj <- strat_results %>%
  mutate(
    logHR = log(HR),
    sig   = ifelse(p.value < 0.05, "*", ""),
    HR_label = ifelse(HR == 0 | is.infinite(CI_high),
                      "N/E", sprintf("%.2f%s", HR, sig)),
    logHR = ifelse(HR_label == "N/E", NA, logHR),
    Gene    = factor(Gene, levels = rev(final_panel)),
    Subtype = factor(Subtype, levels = levels(df$SUBTYPE))
  )

p_adj <- ggplot(hm_adj, aes(x = Subtype, y = Gene, fill = logHR)) +
  geom_tile(color = "white") +
  geom_text(aes(label = HR_label), size = 3.5) +
  scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                       midpoint = 0, na.value = "gray70", name = "log(HR)") +
  labs(title = "Adjusted HR — Gene × Subtype",
       subtitle = "Multivariable Cox (adjusted for age + stage); * = p < 0.05, N/E = not estimable",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank())

print(p_adj)
ggsave(file.path(plot_dir, "HR_heatmap_adjusted.png"),
       p_adj, width = 8, height = 6, dpi = 300)
```

```{r forest-by-subtype, fig.width=11, fig.height=9}
forest_df <- strat_results %>%
  filter(HR > 0, !is.infinite(CI_high), !is.na(HR)) %>%
  mutate(Gene = factor(Gene, levels = rev(final_panel)),
         sig  = ifelse(p.value < 0.05, "Significant", "Not significant"))

p_forest <- ggplot(forest_df, aes(x = HR, y = Gene, color = sig)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray50") +
  geom_errorbarh(aes(xmin = CI_low, xmax = CI_high),
                 height = 0.25, linewidth = 0.8) +
  geom_point(size = 3) +
  scale_color_manual(values = c("Significant"     = "#C0392B",
                                "Not significant" = "#7F8C8D"),
                     name = NULL) +
  facet_wrap(~ Subtype, scales = "free_x", ncol = 2) +
  scale_x_log10() +
  labs(title    = "Forest Plot — Adjusted HR by Subtype",
       subtitle = "Multivariable Cox per subtype, adjusted for age + stage (N/E combinations excluded)",
       x = "Hazard Ratio (log scale)", y = NULL) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom",
        strip.background = element_rect(fill = "#003366", color = NA),
        strip.text       = element_text(color = "white", face = "bold"))

print(p_forest)
ggsave(file.path(plot_dir, "Forest_bySubtype.png"),
       p_forest, width = 11, height = 9, dpi = 300)
```

## Multiple testing correction (BH)

```{r bh-correction}
strat_results <- strat_results %>%
  mutate(q_BH    = signif(p.adjust(p.value, method = "BH"), 4),
         sig_raw = ifelse(p.value < 0.05, "*", ""),
         sig_BH  = ifelse(q_BH    < 0.05, "*", ""))

kable(strat_results %>% arrange(q_BH),
      caption = "Subtype-stratified HR with BH-adjusted q-values (32 tests)")

n_raw <- sum(strat_results$p.value < 0.05)
n_bh  <- sum(strat_results$q_BH    < 0.05)
cat("Significant (raw p < 0.05) :", n_raw, "/", nrow(strat_results), "\n")
cat("Significant (BH q < 0.05)  :", n_bh,  "/", nrow(strat_results), "\n\n")

kable(strat_results %>% filter(q_BH < 0.05) %>% arrange(q_BH),
      caption = "Gene × Subtype combinations surviving BH (q < 0.05)")
```

## Proportional hazards check (Schoenfeld) for BH-significant hits

```{r ph-check}
hits <- data.frame(
  Gene    = c("ESR1", "TP53", "CDH1", "TP53"),
  Subtype = c("LumA", "LumA", "LumB", "TNBC")
)

ph_hits <- purrr::pmap_dfr(hits, function(Gene, Subtype) {
  fit <- strat_list[[Subtype]]$fit
  zph <- cox.zph(fit)
  row <- as.data.frame(zph$table) %>%
    tibble::rownames_to_column("variable") %>%
    filter(variable == Gene)
  data.frame(
    Gene      = Gene,
    Subtype   = Subtype,
    chisq     = round(row$chisq, 3),
    p_value   = signif(row$p, 4),
    PH_status = ifelse(row$p < 0.05, "Violated", "Not violated")
  )
})

kable(ph_hits,
      caption = "PH assumption check for the 4 BH-significant gene × subtype hits")
```

## RMST validation (τ = 60 months)

```{r rmst}
tau_months <- 60

rmst_bh <- purrr::pmap_dfr(hits, function(Gene, Subtype) {
  sub <- df %>% filter(SUBTYPE == Subtype)
  arm <- ifelse(sub[[Gene]] == "Mutated", 1L, 0L)
  r <- rmst2(time = sub$OS_MONTHS, status = sub$OS_EVENT,
             arm = arm, tau = tau_months)
  d <- r$unadjusted.result["RMST (arm=1)-(arm=0)", ]
  data.frame(
    Gene, Subtype,
    RMST_diff = round(d["Est."],     2),
    CI_low    = round(d["lower .95"], 2),
    CI_high   = round(d["upper .95"], 2),
    p_value   = signif(d["p"],       4)
  )
})

kable(rmst_bh,
      caption = paste0("RMST validation of BH-significant hits (τ = ",
                       tau_months, " months)"))
```

```{r rmst-forest, fig.width=8, fig.height=4}
plot_df_rmst <- rmst_bh %>%
  mutate(label = paste(Gene, Subtype, sep = " × "))

p_rmst <- ggplot(plot_df_rmst,
                 aes(x = RMST_diff, y = label)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_errorbarh(aes(xmin = CI_low, xmax = CI_high),
                 height = 0.2, linewidth = 0.7, color = "#003366") +
  geom_point(size = 3.5, color = "#003366") +
  geom_text(aes(label = sprintf("%.2f", RMST_diff)),
            vjust = -0.8, size = 3.5) +
  labs(title = "RMST Difference (Mutated − Wild-type, τ = 60 months)",
       subtitle = "Negative = fewer life-months for mutation carriers",
       x = "RMST Difference (months)", y = NULL) +
  theme_classic(base_size = 12)

print(p_rmst)
ggsave(file.path(plot_dir, "RMST_forest.png"),
       p_rmst, width = 8, height = 4, dpi = 300)
```

## Sensitivity analysis — TMB adjustment

```{r tmb-sensitivity}
df_tmb <- df %>%
  filter(!is.na(TMB)) %>%
  mutate(log_TMB = log(TMB + 1))

cat("TMB-adjustable cohort size:", nrow(df_tmb),
    "(", round(nrow(df_tmb) / nrow(df) * 100, 1), "% of full)\n\n")

strat_tmb <- lapply(levels(df_tmb$SUBTYPE), function(st) {
  sub  <- df_tmb %>% filter(SUBTYPE == st)
  form <- as.formula(paste0("Surv(OS_MONTHS, OS_EVENT) ~ ",
                            gene_terms, " + AGE + STAGE + log_TMB"))
  fit  <- suppressWarnings(coxph(form, data = sub))
  tidy(fit, exponentiate = TRUE, conf.int = TRUE) %>%
    filter(grepl("Mutated", term)) %>%
    mutate(Subtype = st,
           Gene    = gsub("Mutated$", "", term),
           HR      = round(estimate, 3),
           CI_low  = round(conf.low, 3),
           CI_high = round(conf.high, 3),
           p.value = signif(p.value, 4)) %>%
    select(Gene, Subtype, HR, CI_low, CI_high, p.value)
}) %>% bind_rows()

compare_tmb <- strat_results %>%
  select(Gene, Subtype, HR_main = HR,
         CIlo_main = CI_low, CIhi_main = CI_high, p_main = p.value) %>%
  left_join(strat_tmb %>%
              select(Gene, Subtype, HR_tmb = HR,
                     CIlo_tmb = CI_low, CIhi_tmb = CI_high, p_tmb = p.value),
            by = c("Gene", "Subtype"))

kable(compare_tmb %>% semi_join(hits, by = c("Gene", "Subtype")),
      caption = "Main vs TMB-adjusted Cox (BH-significant hits only)")
```

```{r tmb-dumbbell, fig.width=8, fig.height=6}
bh_hits <- c("ESR1_LumA", "TP53_LumA", "CDH1_LumB", "TP53_TNBC")

plot_df_tmb <- compare_tmb %>%
  mutate(id = paste(Gene, Subtype, sep = "_")) %>%
  filter(id %in% bh_hits) %>%
  pivot_longer(
    cols = c(HR_main, HR_tmb, CIlo_main, CIlo_tmb, CIhi_main, CIhi_tmb),
    names_to = c(".value", "Analysis"),
    names_pattern = "(HR|CIlo|CIhi)_(main|tmb)"
  ) %>%
  mutate(Analysis = factor(Analysis,
                           levels = c("main", "tmb"),
                           labels = c("Main", "TMB-adjusted")),
         label = paste(Gene, Subtype, sep = " × "))

p_tmb <- ggplot(plot_df_tmb, aes(x = HR, y = label, color = Analysis)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray50") +
  geom_errorbarh(aes(xmin = CIlo, xmax = CIhi),
                 height = 0.2, linewidth = 0.7,
                 position = position_dodge(width = 0.4)) +
  geom_point(size = 3.5, position = position_dodge(width = 0.4)) +
  scale_color_manual(values = c("Main" = "#003366", "TMB-adjusted" = "#16A085")) +
  scale_x_log10(
    breaks = c(1, 1.5, 2, 2.5, 3, 3.5, 4, 5),
    labels = c("1", "1.5", "2", "2.5", "3", "3.5", "4", "5")
  ) +
  labs(title = "Sensitivity — BH-significant results: Main vs TMB-adjusted",
       subtitle = "Error bars = 95% CI",
       x = "Hazard Ratio (log scale)", y = NULL) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")

print(p_tmb)
ggsave(file.path(plot_dir, "Sens_tmb_dumbbell_CI.png"),
       p_tmb, width = 8, height = 6, dpi = 300)
```

## Integrated summary — 4 BH-significant hits

```{r final-summary}
final_summary <- strat_results %>%
  semi_join(hits, by = c("Gene", "Subtype")) %>%
  select(Gene, Subtype,
         HR, CI_low, CI_high, p_raw = p.value, q_BH) %>%
  left_join(ph_hits %>% select(Gene, Subtype, PH_status),
            by = c("Gene", "Subtype")) %>%
  left_join(rmst_bh %>% select(Gene, Subtype,
                                RMST_months = RMST_diff,
                                RMST_p = p_value),
            by = c("Gene", "Subtype")) %>%
  left_join(compare_tmb %>% select(Gene, Subtype, HR_tmb),
            by = c("Gene", "Subtype")) %>%
  arrange(q_BH)

kable(final_summary,
      caption = "Integrated summary of 4 BH-significant gene × subtype hits with all validations")
```

------------------------------------------------------------------------

# Session info

```{r sessioninfo}
sessionInfo()
```
