---
title: "Gene Panel Selection for Final Project"
author: "Yiqiao Zhu"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Objective

Construct a curated breast cancer driver gene panel for the subtype-stratified
survival analysis, by intersecting:

1. **COSMIC Cancer Gene Census (v103)** — externally curated cancer driver genes
   annotated with tumor type (breast-related).
2. **MSK-CHORD 2024 breast cohort** — genes with patient-level mutation rate
   above a pre-specified frequency threshold (ensures adequate statistical
   power in subtype-stratified analyses).

The final panel is the **intersection** of these two lists.

> **Rationale for avoiding OS-based selection:**
> Any gene selection that uses survival outcome (e.g., pooled Lasso-Cox) biases
> the downstream subtype analysis by favoring genes with consistent effects
> across subtypes, and missing subtype-specific heterogeneity — which is
> precisely what this project aims to discover.

---

# Setup

```{r libraries}
library(dplyr)
library(tidyr)
library(readr)
library(knitr)
```

```{r paths}
# ---- File paths (Mac absolute paths; change as needed) --------------------
proj_dir    <- "/Users/yiqiaozhu/Desktop/BIOS 629"
final_dir   <- file.path(proj_dir, "Final Project")
mskchord_dir<- file.path(proj_dir, "msk_chord_2024")

cosmic_path <- file.path(
  final_dir,
  "Cosmic_CancerGeneCensus_Tsv_v103_GRCh38",
  "Cosmic_CancerGeneCensus_v103_GRCh38.tsv"
)

patient_path <- file.path(mskchord_dir, "data_clinical_patient.txt")
sample_path  <- file.path(mskchord_dir, "data_clinical_sample.txt")
mut_path     <- file.path(mskchord_dir, "data_mutations.txt")
```

---

# Step 1 — Load COSMIC Cancer Gene Census

```{r load-cosmic}
cgc <- read.delim(cosmic_path, sep = "\t", header = TRUE,
                  stringsAsFactors = FALSE, quote = "")

cat("CGC total genes:", nrow(cgc), "\n")
cat("Column names:\n"); print(names(cgc))
```

Inspect tier and role distributions:

```{r cosmic-distribution}
cat("\nTIER distribution (Tier 1 = strong evidence, Tier 2 = emerging):\n")
print(table(cgc$TIER, useNA = "ifany"))

cat("\nFirst few rows:\n")
head(cgc %>% select(GENE_SYMBOL, TIER, ROLE_IN_CANCER, TUMOUR_TYPES_SOMATIC), 10)
```

---

# Step 2 — Filter COSMIC for Breast Cancer Drivers

We retain genes whose `TUMOUR_TYPES_SOMATIC` field contains "breast"
(case-insensitive).

```{r filter-breast}
breast_drivers <- cgc %>%
  filter(grepl("breast", TUMOUR_TYPES_SOMATIC, ignore.case = TRUE)) %>%
  select(GENE_SYMBOL, TIER, ROLE_IN_CANCER, TUMOUR_TYPES_SOMATIC)

cat("Breast-related drivers in CGC:", nrow(breast_drivers), "\n")
```

Show the full breast driver list:

```{r show-breast-drivers}
kable(breast_drivers, caption = "COSMIC breast cancer drivers (v103)")
```

Optional: restrict to Tier 1 only (stronger evidence):

```{r breast-tier1}
breast_drivers_tier1 <- breast_drivers %>% filter(TIER == "1")
cat("Breast Tier-1 drivers only:", nrow(breast_drivers_tier1), "\n")
```

---

# Step 3 — Compute Gene Mutation Frequencies in MSK-CHORD BC Cohort

```{r load-msk}
pat  <- read.delim(patient_path, comment.char = "#",
                   stringsAsFactors = FALSE, check.names = FALSE)
samp <- read.delim(sample_path,  comment.char = "#",
                   stringsAsFactors = FALSE, check.names = FALSE)
mut  <- read.delim(mut_path,     comment.char = "#",
                   stringsAsFactors = FALSE, check.names = FALSE)

bc_samp <- samp %>% filter(CANCER_TYPE == "Breast Cancer")
bc_ids  <- unique(bc_samp$PATIENT_ID)

cat("BC samples:", nrow(bc_samp), "\n")
cat("BC patients (unique):", length(bc_ids), "\n")
```

Aggregate mutations to the patient level (a patient is "mutated" in a gene if
**any** of their samples harbors a mutation in that gene):

```{r patient-level-mutation}
bc_mut <- mut %>%
  filter(Tumor_Sample_Barcode %in% bc_samp$SAMPLE_ID) %>%
  inner_join(bc_samp %>% select(SAMPLE_ID, PATIENT_ID),
             by = c("Tumor_Sample_Barcode" = "SAMPLE_ID"))

gene_freq <- bc_mut %>%
  distinct(PATIENT_ID, Hugo_Symbol) %>%
  count(Hugo_Symbol, name = "n_mut") %>%
  mutate(rate_pct = round(n_mut / length(bc_ids) * 100, 2)) %>%
  arrange(desc(n_mut))

cat("Distinct mutated genes in BC cohort:", nrow(gene_freq), "\n\n")
head(gene_freq, 30)
```

---

# Step 4 — Apply Frequency Threshold

Choose a patient-level mutation rate threshold. Default: **≥ 5%**.

```{r freq-threshold}
freq_threshold <- 5    # in percent; adjustable

msk_freq_panel <- gene_freq %>%
  filter(rate_pct >= freq_threshold) %>%
  pull(Hugo_Symbol)

cat("Threshold: mutation rate >=", freq_threshold, "%\n")
cat("MSK BC genes passing threshold:", length(msk_freq_panel), "\n")
print(msk_freq_panel)
```

---

# Step 5 — Intersect COSMIC × MSK Frequency Panel

```{r intersection}
final_panel <- intersect(breast_drivers$GENE_SYMBOL, msk_freq_panel)

cat("==== FINAL PANEL (COSMIC breast drivers ∩ MSK freq >=",
    freq_threshold, "%) ====\n")
cat("N =", length(final_panel), "genes\n\n")
print(final_panel)
```

Cross-reference: which genes were in MSK panel but NOT in COSMIC breast list?

```{r msk-only}
msk_only <- setdiff(msk_freq_panel, breast_drivers$GENE_SYMBOL)
cat("In MSK freq panel but NOT in COSMIC breast list:\n")
print(msk_only)
```

Cross-reference: which COSMIC breast drivers were NOT frequent enough in MSK?

```{r cosmic-only}
cosmic_only <- setdiff(breast_drivers$GENE_SYMBOL, msk_freq_panel)
cat("COSMIC breast drivers with mutation rate <", freq_threshold,
    "% in MSK cohort (excluded):\n")
print(cosmic_only)
```

---

# Step 6 — Final Panel Summary Table

```{r panel-summary}
panel_summary <- gene_freq %>%
  filter(Hugo_Symbol %in% final_panel) %>%
  left_join(
    breast_drivers %>% select(GENE_SYMBOL, TIER, ROLE_IN_CANCER),
    by = c("Hugo_Symbol" = "GENE_SYMBOL")
  ) %>%
  arrange(desc(rate_pct))

kable(panel_summary,
      caption = "Final gene panel with mutation frequency and COSMIC role")
```

---

# Step 7 — Save Outputs

Save the final panel + summary for downstream analysis scripts.

```{r save}
# ---- Create a dataset subfolder under Final Project --------------------
dataset_dir <- file.path(final_dir, "dataset")
dir.create(dataset_dir, showWarnings = FALSE, recursive = TRUE)

# Final panel as a character vector saved to RDS (simple to reload)
saveRDS(final_panel, file.path(dataset_dir, "final_panel.rds"))

# Summary table as CSV
write.csv(panel_summary,
          file.path(dataset_dir, "final_panel_summary.csv"),
          row.names = FALSE)

# Also save breast drivers list + gene frequency table for reference
write.csv(breast_drivers,
          file.path(dataset_dir, "cosmic_breast_drivers.csv"),
          row.names = FALSE)

write.csv(gene_freq,
          file.path(dataset_dir, "msk_bc_gene_frequency.csv"),
          row.names = FALSE)

cat("Saved to:", dataset_dir, "\n")
cat(" - final_panel.rds\n")
cat(" - final_panel_summary.csv\n")
cat(" - cosmic_breast_drivers.csv\n")
cat(" - msk_bc_gene_frequency.csv\n")
```

---

# Notes / Next Steps

- If the resulting panel is too small (< 8 genes), consider lowering the
  frequency threshold to 2% (`freq_threshold <- 2`), or relaxing the COSMIC
  breast filter (e.g., include any gene whose `TISSUE_TYPE` field mentions
  breast-related tissues).
- If the panel feels too large (> 20 genes), restrict COSMIC to **Tier 1**
  only using `breast_drivers_tier1` from Step 2.
- The resulting `final_panel` will be loaded by the downstream analysis
  scripts (descriptive stats, KM, Cox, interaction, RMST, sensitivity
  analyses).
